Transitivity of Subtyping for Intersection Types

### 2020-05-19 papers/LMCS-subtyping-intersections/response.txt

>> 1. lack of details for the motivating connection to Siek's broader
>>    project(s), alluded to in the Introduction but too briefly to
>>    assess the value of this work on that basis.

The revision includes a new section "Motivations from Verified
Compilation" to address this problem.

>> 2. failure, as also in the Laurent paper that seems to have been the
>>    starting point for this work, to consider the deep line of work on
>>    logical treatments of subtyping.
>>
>> ...  In contrast, Siek's system handles function types through some
>> very particular meta-level operations on type expressions and side
>> conditions to the function subtyping rule (->').  I worry that
>> these conditions will be brittle, and not work well in a setting
>> where compositionality at the type level is required.  Imagine, for
>> example, allowing type-level functions as in System F-omega.  Then
>> particular checks on the forms of type expressions will fail in the
>> presence of applications of type variables (i.e., abstracted type
>> expressions), since we will not have a concrete type expression to
>> apply these various checks and meta-level operations to.  So an
>> approach that is based just on syntactic conditions on type
>> expressions seems fragile and not likely to be extensible to richer
>> type systems.

Yes, the logical treatments of subtyping are useful for extending to
richer type systems for programming languages. However, the motivation
for the present work is in constructing filter models (denotational
semantics) for programming languages. For that purpose, there is no
need for richer type systems such as System F-omega; intersection
types already provide what is needed to express the runtime behavior
of functional programs. The revision includes changes to the
introduction to differentiate the varied uses of intersection type
systems and to make it clear that the present article's motivation
comes from the realm of filter models, not type systems for
programming languages.

@@

I've changed the notation for \in and \subseteq to make it clear that
these notions are different from the usual ones and added some
explanation to provide an intuition for what they accomplish.

Reviewer 3
==========

>> The only question I had is specifically related to the recent adoption
>> of intersection types in Scala and in formal work on the underlying
>> DOT calculus. The need for transitivity elimination is also described
>> in the OOPSLA'16 paper "Type Soundness for Dependent Object Types",
>> and the apparent solution adopted there was to "push back" top-level
>> uses of the transitivity rule further up into derivations, based on a
>> lemma "A <: B /\ B <! C ==> A <! C" where "A <! C" is a subtyping
>> derivation that does not end in the transitivity rule but may use it
>> internally. While it is not possible to get rid of transitivity
>> entirely in DOT, it seems possible to apply the "pushback" process to
>> exhaustion in the BCD setting considered in this paper, and in fact
>> I'm wondering if this is roughly equivalent to the induction metric
>> "<<". Perhaps the author can give this some thought, and maybe a
>> discussion of this piece of related work would be a useful addition
>> (if only to strengthen the motivation).

That is an interesting connection. I think that this style of proof
would work using the new function rule (→′) but not with the BCD rules
(→) and (→∩). I've added a discussion of this connection to the
revision.

@@

but then he was forced to add another rule

C ≤ A → D   D ≤ B
-----------------
C ≤ A → B

that does not satisfy the subformula property.

>> Minor notes:
>> The paper has a few formatting issues with text running into the
>> margin (p1, p8, some URLs in the bibliography).
>> Typos: 
>> "and then then prove transitivity" (p2) double "then"
>> "Before moving on, we make note of some basic facts" perhaps "take
>> note" or "make a note of"?
>> Bibliography:
>> "M. Dezani-Ciancaglini, F. Honsell, and Y. Motohama. Compositional
>> characterisations of -terms" missing a lambda symbol
>> "E. Engeler. Algebras and combinators. algebra universalis" capitalize
>> journal title (?)
>> "H. Ishihara and T. Kurata. Completeness of intersection and union
>> type assignment systems for call-by-value -models." missing lambda

### 2020-05-19 papers/LMCS-subtyping-intersections/paper.tex
check

\documentclass{article}
%\usepackage{cmbright}
\usepackage{natbib}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{mathabx}
\usepackage{stmaryrd}
\usepackage{semantic}
%\usepackage{fullpage}
%\usepackage{fontspec,unicode-math}
\usepackage{hyperref}

@@

\begin{abstract}
  The subtyping rules for intersection types traditionally employ a
  transitivity rule (Barendregt et al. 1983), which means that
  subtyping does not satisfy the subformula property, making it more
  difficult to use in filter models for compiler verification.
  Laurent develops a sequent-style subtyping system, without
  transitivity, and proves transitivity via a sequence of six lemmas
  that culminate in cut-elimination (2018). This article develops a
  subtyping system in regular style that omits transitivity and
  provides a direct proof of transitivity, significantly reducing the
  length of the proof, exchanging the six lemmas for just
  one. Inspired by Laurent's system, the rule for function types is
  essentially the $\beta$-soundness property.  The new system
  satisfies the ``subformula conjunction property'': every type
  occuring in the derivation of $A <: B$ is a subformula of $A$ or
  $B$, or an intersection of such subformulas. The article proves that
  the new subtyping system is equivalent to that of Barendregt, Coppo,
  and Dezani-Ciancaglini.
\end{abstract}

@@

Intersection types were invented by Coppo, Dezani-Ciancaglini, and
Salle to study normalization in the lambda
calculus~\citep{Coppo:1979aa}. Subsequently intersection types have
been used for at least three purposes:
\begin{enumerate}
  \item in type systems for programming
    languages~\citep{Reynolds:1988aa,Pierce:1991aa,Castagna:2014aa,Chaudhuri:2014aa,Oliveira:2016aa,Amin:2017aa,Muehlboeck:2018aa,Bi:2019aa,Dunfield:2019aa,Microsoft:TypeScript2020aa},
  \item in precise static
    analyses~\citep{Turbak:1997aa,Palsberg:1998aa,Mossin:2003aa,Simoes:2007aa},
    and
  \item in filter models for the denotational semantics of a variety
    of lambda
    calculi~\citep{Coppo:1980ab,Coppo:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Honsell:1999aa,Ishihara:2002aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa,Barendregt:2013aa}.
\end{enumerate}
The motivation for this article comes from the third use of
intersection types, in denotational semantics.  We are interested in
constructing filter models for functional programming languages in
support of mechanized compiler
verification. Section~\ref{sec:motivation} describes how that research
motivates the results in this article.

@@

Our interest in intersection type systems stems from our work on
mechanized correctness proofs for compilers for functional programming
languages. We are exploring whether denotational semantics can
streamline such proofs compared to the operational techniques such as
those used to prove the correctness of
CakeML~\citep{Kumar:2014aa,Owens:2017aa}. In particular, we are
constructing filter models (based on intersection type systems)
because they are relatively straightforward to mechanize in a proof
assistant compared to traditional domain
theory~\citep{Benton:2009ab,Dockins:2014aa}.  In a filter model, the
meaning of a program $M$ is the set of all types that can be assigned
to it by an intersection type system:
\[
   \llbracket M \rrbracket = \{ A \mid \emptyset \vdash M : A \}
\]
Most type systems are much too coarse of an abstraction to be used as
filter models. Intersection type systems are special in that they can
completely characterize the runtime behavior of a program.  To catch a
glimpse of how filter models work, consider the meaning of the
identity function in the $\lambda$-calculus.
\[
  \llbracket \lambda x.\, x \rrbracket =
      \{ c_0 \to c_0, c_1 \to c_1, c_2 \to c_2, \ldots \}
\]
If one takes each type constant $c_i$ to represent a singleton type
for the number $i$, then the set of types assigned to the identity
function starts to look like the graph of the function. Indeed, filter
models are closely related to graph
models~\citep{Scott:1976lq,Engeler:1981aa,Plotkin:1993ab,Barendregt:2013aa}.

@@

This problem is not conceptually difficult to solve; one just needs to
restrict the graphs to only allow approximations of functions. This
can be accomplished by requiring every entry in a function's graph to
be ``consistent''. Two entries are consistent, written $(A \to B) \sim
(A' \to B')$, when either $A \not\sim A'$ or $A \sim A'$ and $B \sim
B'$. Two type constants are consistent when they are equal. One would
then hope to prove that for any term $M$, if $\Gamma \vdash M : C$ and
$\Gamma \vdash M : C'$, then $C \sim C'$ as long as $C$, $C'$, and
$\Gamma$ are self consistent. A key lemma needed for that proof is
that consistency is upward closed with respect to subtyping: if $A
\sim B$, $A <: C$, $B <: D$, then $C \sim D$.

This brings us back to the transitivity rule and the subformula
property. The proof of the above lemma hits a snag in the cases where
$A <: C$ and $B <: D$ are derived using the transitivity rule. The
intermediate type is not guaranteed to be self consistent. We explored
adding this guarantee to the transitivity rule, but it caused a
significant increase in obligations elsewhere in our proofs. By some
serendipity, around that moment in time \citet{Laurent:2018aa}
published the first subtyping system for intersection types that
removes the transitivity rule and satisfies the subformula property.
Laurent developed a sequent-style subtyping system, written $\Gamma
\vdash B$, where $\Gamma$ is a sequence of types $A_1,\ldots,A_n$. The
intuition is that $A_1,\ldots,A_n \vdash B$ corresponds to $A_1 \cap
\cdots \cap A_n \leq B$. This system satisfies the subformula property
and is equivalent to the BCD system. To prove this, Laurent
establishes six lemmas that culminate in cut-elimination, from which
transitivity follows. Laurent mechanized these proofs in Coq.

We immediately took advantage of Laurent's result to complete a filter
model for the ISWIM
language~\citep{Landin:1966la,G.-D.-Plotkin:1975on,Felleisen:2009aa}.
We mechanized the formal development and proofs using the Isabelle
proof assistant~\citep{Siek:2018aa}.

@@

The present article discusses how we were able to import the key
feature from Laurent's system, the rule for subtyping between function
types, into the BCD subtyping system, replacing the function $(\to)$,
distributivity $({\to}{\cap})$, and transitivity (trans) rules.  We
then prove transitivity without the sequence of six lemmas, instead
using one lemma that has been used previously to prove the inversion
principle for subtyping of function
types~\citep{Barendregt:2013aa}. The new subtyping system does not
technically satisfy the subformula property but it satisfies what we
call the \emph{subformula conjunction property}: every type occuring
in the derivation of $A <: B$ is a subformula of $A$ or $B$ or an
intersection of such subformulas.  This property is enough to ensure
that if $A$ and $B$ are self consistent, then so are all the types
that appear in the derivation of $A <: B$.  Thus, the new subtyping
system is suitable for mechanizing filter models.

@@

The rest of this article is organized as follows. We discuss two
recent developments regarding subtyping for intersection types in
Section~\ref{sec:recent-developments}.  We present our new subtyping
system in Section~\ref{sec:new-subtyping}, prove transitivity in
Section~\ref{sec:trans}, and prove its equivalence to BCD subtyping in
Section~\ref{sec:equiv}. We conclude the artcle in
Section~\ref{sec:conclude}.

The definitions and results in this article have been machine checked
in Agda, in the file \texttt{agda/TransSubInter.agda} in the following
repository.

\begin{center}
  \url{https://github.com/jsiek/denotational_semantics}
\end{center}

@@

\citet{Bi:2018aa} present a subtyping algorithm that satisfies the
subformula property and proves that it is equivalent to BCD
subtyping. Their system is based on the decision procedure
of~\citet{Pierce:1989aa} and takes the form $\mathcal{L} \vdash A
\prec: B$, where $\mathcal{L}$ is a queue of types that are peeled off
from the domain of the type on the right. So the judgement
$\mathcal{L} \vdash A \prec: B$ is equivalent to $A \leq \mathcal{L}
\to B$ where $\mathcal{L} \to B$ is defined as
\begin{align*}
  [] \to B &= B\\
  (\mathcal{L},A) \to B &= \mathcal{L} \to (A \to B)
\end{align*}
The proof of transitivity for this system is an adaptation of Pierce's
but it corrects some errors and introduces more lemmas concerning an
auxilliary notion called reflexive supertypes.

@@

Our new subtyping system relies on a few definitions that are given in
Figure~\ref{fig:aux}. These include the partial functions $\dom{A}$
and $\cod{A}$, the $\topP{A}$ and $\topInCod{A}$ predicates, and the
relations $A \inside B$ and $A \containedin B$.
%
The $\dom{A}$ and $\cod{A}$ partial functions return the domain or
codomain if $A$ is a function type, respectively. If $A$ is an
intersection $A_1 \cap A_2$, then $\dom{A}$ is the intersection of the
domain of $A_1$ and $A_2$.  If $A$ is an atom, $\dom{A}$ is
undefined. Likewise for $\cod{A}$. For example, if $A = (A_1 \to B_1)
\cap \cdots \cap (A_n \to B_n)$, then $\dom{A} = A_1 \cap \cdots \cap
A_n$ and $\cod{A} = B_1 \cap \cdots \cap B_n$.  When $\dom{A}$ or
$\cod{A}$ appears in lemma or theorem statement, we implicitly assume
that $A$ is a type such that $\dom{A}$ and $\cod{A}$ are defined.
%
The $\topP{A}$ predicate identifies types that are equivalent to
$\TOP$. (See Propostion~\ref{prop:top}(\ref{prop:AllBot-⊑-any}).) The
$\topInCod{A}$ predicate identifies types that have $\TOP$ in their
codomain.
%
The relations $A \inside B$ and $A \containedin B$ enable the
treatment of a sequence of intersections $B_1 \cap \cdots \cap B_n$
(modulo associativity and commutativity) as if it were a set of types
$\{ B_1, \ldots, B_n \}$, where each $B_i$ is an atom or a function
type.  We say that $A$ is a part of $B$ if $A \inside B$ and we say
that $B$ \emph{contains} $A$ if $A \containedin B$.

@@

The new intersection subtyping system, with judgements of the form $A
<: B$, is defined in Figure~\ref{fig:new-subtyping}. First, it does
not include the (trans) rule.  It also replaces the (refl) rule with
reflexivity for atoms (refl$_\alpha$). The most important rule is the
one for function types $(\to')$, which subsumes $(\to)$ and
$({\to}{\cap})$ in BCD subtyping.  The $(\to')$ rule essentially turns
the $\beta$-soundness property into a subtyping
rule~\citep{Barendregt:2013aa}. The $(\to')$ rule says that a type $A$
is a subtype of a function type $C \to D$ if a type contained in $A$,
call it $B$, has domain and codomain that are larger and smaller than
$C$ and $D$, respectively. The use of containment enables this rule to
absorb uses of (incl$_L$) and (incl$_R$) on the left.  The side
conditions $\neg\;\topP{B}$ and $\neg\;\topInCod{D}$ are needed
because of the $(\TOP{\to}')$ rule, which in turn is needed to
preserve types under $\eta$-reduction.  In the many systems that do
not involve $\eta$-reduction, the $(\TOP{\to}')$ rule can be omitted,
as well as these side conditions. The rules (lb$_L$) and (lb$_R$)
adapt (incl$_L$) and (incl$_R$) to a system without transitivity, and
have appeared many times in the literature~\citep{Bakel:1995aa}.  The
$(\TOP{\to}')$ rule generalizes the $(\TOP{\to})$ rule, replacing the
$\TOP$ on the left with any type $A$, because for transitivity, any
type is below $\TOP$. The $(\TOP{\to}')$ rule replaces the $\TOP$ in
the codomain on the right with any type $D$ that is equivalent to
$\TOP$.

@@

\bibliographystyle{abbrvnat}
\bibliography{all}

@@

% LocalWords:  dom subtyping Barendregt et al subformula sequent aa
% LocalWords:  BCD lclr refl glb topInCod Coppo Dezani Ciancaglini fk
% LocalWords:  Salle Engeler Honsell Abramsky Plotkin Rocca Alessi
% LocalWords:  subsumption subtype denotational ISWIM Landin Ishihara
% LocalWords:  Felleisen Muehlboeck ISC Agda codomain reflexivity
% LocalWords:  Bakel

>>>>

\documentclass{article}
%\usepackage{cmbright}
\usepackage[numbers]{natbib}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{mathabx}
\usepackage{stmaryrd}
\usepackage{semantic}
%\usepackage{fullpage}
%\usepackage{fontspec,unicode-math}
\usepackage{hyperref}

@@

\begin{abstract}
The subtyping rules for intersection types traditionally employ a
transitivity rule (Barendregt et al. 1983), which means that
subtyping does not satisfy the subformula property, making it more
difficult to use in filter models for compiler verification.
Laurent develops a sequent-style subtyping system, without
transitivity, and proves transitivity via a sequence of six lemmas
that culminate in cut-elimination (2018). This article develops a
subtyping system in regular style that omits transitivity and
provides a direct proof of transitivity, significantly reducing the
length of the proof, exchanging the six lemmas for just
one. Inspired by Laurent's system, the rule for function types is
essentially the $\beta$-soundness property.  The new system
satisfies the ``subformula conjunction property'': every type
occurring in the derivation of $A <: B$ is a subformula of $A$ or
$B$, or an intersection of such subformulas. The article proves that
the new subtyping system is equivalent to that of Barendregt, Coppo,
and Dezani-Ciancaglini.
\end{abstract}

@@

Intersection types were invented by Coppo, Dezani-Ciancaglini, and
Salle to study normalization in the lambda
calculus~\citep{Coppo:1979aa}. Subsequently intersection types have
been used for at least three purposes:
\begin{enumerate}
  \item in type systems~\citep{Reynolds:1988aa,Pierce:1991aa,Castagna:2014aa,Chaudhuri:2014aa,Oliveira:2016aa,Amin:2017aa,Muehlboeck:2018aa,Bi:2019aa,Dunfield:2019aa,Microsoft:TypeScript2020aa},
  \item in precise static
    analyses~\citep{Turbak:1997aa,Palsberg:1998aa,Mossin:2003aa,Simoes:2007aa},
    and
  \item in the denotational semantics for a variety of lambda
    calculi~\citep{Coppo:1980ab,Coppo:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Honsell:1999aa,Ishihara:2002aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa,Barendregt:2013aa}.
\end{enumerate}
The motivation for this article comes from the third use of
intersection types, in denotational semantics. We are interested in
constructing filter models for the denotational semantics of
functional programming languages in support of mechanized compiler
verification. Section~\ref{sec:motivation} describes how that research
motivates the results in this article.

@@

Our interest in intersection type systems stems from our work on
mechanized correctness proofs for compilers for functional programming
languages. We are exploring whether denotational semantics can
streamline such proofs compared to the operational techniques such as
those used to prove the correctness of
CakeML~\citep{Kumar:2014aa,Owens:2017aa}. In particular, we are
constructing filter models because they are relatively straightforward
to mechanize in a proof assistant compared to traditional domain
theory~\citep{Benton:2009ab,Dockins:2014aa}.  In a filter model, the
meaning of a program $M$ is the set of all types that can be assigned
to it by an intersection type system:
\[
   \llbracket M \rrbracket = \{ A \mid \emptyset \vdash M : A \}
\]
Most type systems are much too coarse of an abstraction to be used as
filter models. Intersection type systems are special in that they can
completely characterize the runtime behavior of a program.  To catch a
glimpse of how filter models work, consider the meaning of the
identity function in the $\lambda$-calculus.
\[
  \llbracket \lambda x.\, x \rrbracket =
      \{ c_0 \to c_0, c_1 \to c_1, c_2 \to c_2, \ldots \}
\]
If one takes each type constant $c_i$ to represent a singleton type
for the number $i$, then the set of types assigned to the identity
function starts to look like the graph of the function. Indeed, filter
models are closely related to graph
models~\citep{Scott:1976lq,Engeler:1981aa,Plotkin:1993ab,Barendregt:2013aa}.

@@

To solve this problem, one needs to restrict the graphs to only allow
approximations of functions. This can be accomplished by requiring
every entry in a function's graph to be ``consistent''. Two entries
are consistent, written $(A \to B) \sim (A' \to B')$, when either $A
\not\sim A'$ or $A \sim A'$ and $B \sim B'$. Two type constants are
consistent when they are equal. One would then hope to prove that for
any term $M$, if $\Gamma \vdash M : C$ and $\Gamma \vdash M : C'$,
then $C \sim C'$ as long as $C$, $C'$, and $\Gamma$ are self
consistent. A key lemma needed for that proof is that consistency is
upward closed with respect to subtyping: if $A \sim B$, $A <: C$, $B
<: D$, then $C \sim D$.

This brings us back to the transitivity rule and the subformula
property. The proof of the above lemma hits a snag in the cases where
$A <: C$ and $B <: D$ are derived using the transitivity rule. The
intermediate type is not guaranteed to be self consistent. We explored
adding this guarantee to the transitivity rule, but it caused a
significant increase in obligations elsewhere in our proofs. By some
serendipity, around that time \citet{Laurent:2018aa} published a
subtyping system that removes the transitivity rule and satisfies the
subformula property.  Laurent developed a sequent-style subtyping
system, written $\Gamma \vdash B$, where $\Gamma$ is a sequence of
types $A_1,\ldots,A_n$. The intuition is that $A_1,\ldots,A_n \vdash
B$ corresponds to $A_1 \cap \cdots \cap A_n \leq B$. This system
satisfies the subformula property and is equivalent to the BCD
system. To prove this, Laurent establishes six lemmas that culminate
in cut-elimination, from which transitivity follows. Laurent
mechanized these proofs in Coq.

We immediately used Laurent's result to complete a filter model for
ISWIM~\citep{Landin:1966la,G.-D.-Plotkin:1975on,Felleisen:2009aa} and
mechanized the results in the Isabelle proof
assistant~\citep{Siek:2018aa}.

@@

The present article discusses how we were able to import the key
feature from Laurent's system, the rule for subtyping between function
types, into the BCD subtyping system, replacing the function $(\to)$,
distributivity $({\to}{\cap})$, and transitivity (trans) rules.  We
then prove transitivity without the sequence of six lemmas, instead
using one lemma that has been used previously to prove the inversion
principle for subtyping of function
types~\citep{Barendregt:2013aa}. The new subtyping system does not
technically satisfy the subformula property but it satisfies what we
call the \emph{subformula conjunction property}: every type occurring
in the derivation of $A <: B$ is a subformula of $A$ or $B$ or an
intersection of such subformulas.  This property is enough to ensure
that if $A$ and $B$ are self consistent, then so are all the types
that appear in the derivation of $A <: B$.  Thus, the new subtyping
system is suitable for mechanizing filter models.

@@

The rest of this article is organized as follows. We discuss two
recent developments regarding subtyping for intersection types in
Section~\ref{sec:recent-developments}.  We present our new subtyping
system in Section~\ref{sec:new-subtyping}, prove transitivity in
Section~\ref{sec:trans}, and prove its equivalence to BCD subtyping in
Section~\ref{sec:equiv}. We conclude the article in
Section~\ref{sec:conclude}.

The definitions and results in this article have been machine checked
in Agda, in the file \texttt{agda/TransSubInter.agda} in the following
repository.
\begin{center}
  \url{https://github.com/jsiek/denotational_semantics}
\end{center}

@@

\citet{Bi:2018aa} present a subtyping algorithm that satisfies the
subformula property and proves that it is equivalent to BCD
subtyping. Their system is based on the decision procedure
of~\citet{Pierce:1989aa} and takes the form $\mathcal{L} \vdash A
\prec: B$, where $\mathcal{L}$ is a queue of types that are peeled off
from the domain of the type on the right. So the judgment
$\mathcal{L} \vdash A \prec: B$ is equivalent to $A \leq \mathcal{L}
\to B$ where $\mathcal{L} \to B$ is defined as
\begin{align*}
  [] \to B &= B\\
  (\mathcal{L},A) \to B &= \mathcal{L} \to (A \to B)
\end{align*}
The proof of transitivity for this system is an adaptation of Pierce's
but it corrects some errors and introduces more lemmas concerning an
auxiliary notion called reflexive supertypes.

@@

Our new subtyping system relies on a few definitions that are given in
Figure~\ref{fig:aux}. These include the partial functions $\dom{A}$
and $\cod{A}$, the $\topP{A}$ and $\topInCod{A}$ predicates, and the
relations $A \inside B$ and $A \containedin B$.
%
The $\dom{A}$ and $\cod{A}$ partial functions return the domain or
codomain if $A$ is a function type, respectively. If $A$ is an
intersection $A_1 \cap A_2$, then $\dom{A}$ is the intersection of the
domain of $A_1$ and $A_2$.  If $A$ is an atom, $\dom{A}$ is
undefined. Likewise for $\cod{A}$. For example, if $A = (A_1 \to B_1)
\cap \cdots \cap (A_n \to B_n)$, then $\dom{A} = A_1 \cap \cdots \cap
A_n$ and $\cod{A} = B_1 \cap \cdots \cap B_n$.  When $\dom{A}$ or
$\cod{A}$ appears in lemma or theorem statement, we implicitly assume
that $A$ is a type such that $\dom{A}$ and $\cod{A}$ are defined.
%
The $\topP{A}$ predicate identifies types that are equivalent to
$\TOP$. (See Proposition~\ref{prop:top}(\ref{prop:AllBot-⊑-any}).) The
$\topInCod{A}$ predicate identifies types that have $\TOP$ in their
codomain.
%
The relations $A \inside B$ and $A \containedin B$ enable the
treatment of a sequence of intersections $B_1 \cap \cdots \cap B_n$
(modulo associativity and commutativity) as if it were a set of types
$\{ B_1, \ldots, B_n \}$, where each $B_i$ is an atom or a function
type.  We say that $A$ is a part of $B$ if $A \inside B$ and we say
that $B$ \emph{contains} $A$ if $A \containedin B$.

@@

The new intersection subtyping system, with judgments of the form $A
<: B$, is defined in Figure~\ref{fig:new-subtyping}. First, it does
not include the (trans) rule.  It also replaces the (refl) rule with
reflexivity for atoms (refl$_\alpha$). The most important rule is the
one for function types $(\to')$, which subsumes $(\to)$ and
$({\to}{\cap})$ in BCD subtyping.  The $(\to')$ rule essentially turns
the $\beta$-soundness property into a subtyping
rule~\citep{Barendregt:2013aa}. The $(\to')$ rule says that a type $A$
is a subtype of a function type $C \to D$ if a type contained in $A$,
call it $B$, has domain and codomain that are larger and smaller than
$C$ and $D$, respectively. The use of containment enables this rule to
absorb uses of (incl$_L$) and (incl$_R$) on the left.  The side
conditions $\neg\;\topP{B}$ and $\neg\;\topInCod{D}$ are needed
because of the $(\TOP{\to}')$ rule, which in turn is needed to
preserve types under $\eta$-reduction.  In the many systems that do
not involve $\eta$-reduction, the $(\TOP{\to}')$ rule can be omitted,
as well as these side conditions. The rules (lb$_L$) and (lb$_R$)
adapt (incl$_L$) and (incl$_R$) to a system without transitivity, and
have appeared many times in the literature~\citep{Bakel:1995aa}.  The
$(\TOP{\to}')$ rule generalizes the $(\TOP{\to})$ rule, replacing the
$\TOP$ on the left with any type $A$, because for transitivity, any
type is below $\TOP$. The $(\TOP{\to}')$ rule replaces the $\TOP$ in
the codomain on the right with any type $D$ that is equivalent to
$\TOP$.

@@

%\bibliographystyle{abbrvnat}
\bibliographystyle{plainnat}
\bibliography{all}

@@

% LocalWords:  dom subtyping Barendregt et al subformula sequent aa
% LocalWords:  BCD lclr refl glb topInCod Coppo Dezani Ciancaglini fk
% LocalWords:  Salle Engeler Honsell Abramsky Plotkin Rocca Alessi lq
% LocalWords:  subsumption subtype denotational ISWIM Landin Ishihara
% LocalWords:  Felleisen Muehlboeck ISC Agda codomain reflexivity Coq
% LocalWords:  Bakel subformulas Castagna Chaudhuri Oliveira Amin
% LocalWords:  Dunfield TypeScript Turbak Palsberg Mossin Simoes Siek
% LocalWords:  compositional CakeML Kumar Dockins runtime Appel agda
% LocalWords:  subexpression Tarditi denotationally distributivity
% LocalWords:  Scala supertypes Odersky Rompf pushback associativity
% LocalWords:  commutativity
